---
title: "Advent of Code 2020"
author: "Stephanie M Haas"
date: "12/3/2020"
output: 
  html_document:
    theme: yeti
    toc: true
    toc_depth: 3
    toc_float: yes
---


Here are my solutions for advent of code 2020. Advent of code is a set of 25 coding problems released daily that gain progressively harder throughout the month of December. They can be completed in any coding language; however, I am exclusively using R.

Full descriptions of the problems and the data is available at https://adventofcode.com/2020



```{r}
library(tidyverse)
library(gmp) #for big numbers!
```


# Day 1

https://adventofcode.com/2020/day/1



## Part 1

For day 1, part 1 we had a list of numbers and we had to find the two numbers that summed to 2020.

```{r}
day1 <- read_csv("data/day1.csv")
head(day1)

total <- 2020
high_values <- day1 %>% filter(expense > total/2 & expense != total)
low_values <- day1 %>% filter(expense < total/2 & expense != total)
pair <- high_values %>% 
  mutate(search_value = total - expense) %>% 
    inner_join(low_values, by = c("search_value" = "expense"))
solution_day1_p1 <- pair[1,1] * pair[1,2]
```

## Part 2

For part 2 we had to find which 3 numbers summed to 2020


```{r}

first_value <- day1 %>% filter(expense <= 2020/2)

list <- lapply(first_value$expense, function(iter){
  total <- 2020 - iter
  high_values <- day1 %>% filter(expense > total/2 & expense != total)
  low_values <- day1 %>% filter(expense < total/2 & expense != total)
  pair <- high_values %>% 
      mutate(search_value = (expense - total) * -1) %>% 
      inner_join(low_values, by = c("search_value" = "expense"))
  return(iter * pair[1,1] * pair[1,2])
})

solution_day1_p2 <- unique(list)[1]

k <- NA
i <- 1
while(is.na(k)){
  iter <- first_value$expense[i]
  total <- 2020 - iter
  high_values <- day1 %>% filter(expense > total/2 & expense != total)
  low_values <- day1 %>% filter(expense < total/2 & expense != total)
  pair <- high_values %>% 
    mutate(search_value = total - expense) %>% 
    inner_join(low_values, by = c("search_value" = "expense"))
  k = iter * pair[1,1] * pair[1,2]
  i = i + 1
}
solution_day1_p2 <- k

```

# Day 2

https://adventofcode.com/2020/day/2

## Part 1

For day 2 we had a set of passwords with criteria dictating the minimum and maximum number of times a stated letter had to appear in the code. We had to find how many passwords were valid (i.e. followed their criteria).

```{r}
day2 <- read_csv("data/day2.csv")
head(day2)

day2_split <- sapply(day2$value, strsplit, split = ": | |-")
day2_split <- do.call("rbind", day2_split)

colnames(day2_split) <- c("min", "max", "letter", "password")

day2_split <- as_tibble(day2_split) %>% 
  mutate(valid = NA,
         min = as.numeric(min),
         max = as.numeric(max)) %>% 
  rowid_to_column()

errors <- lapply(day2_split$rowid, function(row){
  code <- str_extract_all(day2_split$password[row], "[a-z]")
  code <- do.call("rbind", code)
  iter <- length(code[code == day2_split$letter[row]])
  if (iter >= day2_split$min[row] & iter <= day2_split$max[row]){

    return(1)

  }else{

    return(0)

  }
})

solution_day2_p1 <- do.call("sum", errors)

```


## Part 2

Part 2 was the same as part 1 except instead of min and max we had the specific digit in the code. So each letter had to appear at exactly one of the two given indices.


```{r}

errors <- lapply(day2_split$rowid, function(row){
  code <- str_extract_all(day2_split$password[row], "[a-z]")
  code <- do.call("rbind", code)
  first <- code[day2_split$min[row]]
  second <- code[day2_split$max[row]]
  if (xor(first == day2_split$letter[row], second == day2_split$letter[row])){
    
    return(1)
    
  }else{
    
    return(0)
    
  }
})

solution_day2_p2 <- do.call("sum", errors)


```


# Day 3

https://adventofcode.com/2020/day/3

## Part 1

In Day 3 we had a matrix composed of "." and "#" and given a line that started at the top left corner and progressing down 1 and to the right 3 we had to determine how many #'s intersected with the line. The matrix duplicated itself to the right infinitely until it reached the bottom.

```{r}
day3 <- read_csv("data/day3.csv", col_names = "hill")
head(day3)

day3 <- day3 %>% 
  mutate(hill = gsub("\\.", 0, hill),
         hill = gsub("#", 1, hill))

day3_split <- sapply(day3$hill, str_extract_all, "\\d")
day3_split <- do.call("rbind", day3_split) %>% 
  as_tibble()


max <- dim(day3_split)[1]

tree_num <- lapply(1:max, function(row){
  column <- ((row - 1) * 3 + 1) %% 31
  if (column == 0) {column = 31}
  return(as.numeric(day3_split[row, column]))
})

solution_day3_p1 <- do.call("sum", tree_num)

```

## Part 2

Part 2 extended part 1 except we were to find the product of the number of #'s intersected for several lines including down 1 and right 1, down 1 and right 3, down 1 and right 5, down 1 and right 7, and down 2 and right 1.

```{r}
max <- dim(day3_split)[1]
right <- c(seq(1,7,by = 2), 1)
down <- c(1,1,1,1,2)


slope_sum <- function(row, down, right){
   row_adjust <- (row - 1) * down + 1
    column <- ((row - 1) * right + 1) %% 31
    if (column == 0) {column = 31}
    return(as.numeric(day3_split[row_adjust, column]))
}

nums <- mapply(function(down, right){
  tree_num <- lapply(1:ceiling(max/down), slope_sum, down = down, right = right) 
  return(do.call("sum", tree_num))
}, down, right)

solution_day3_p2 <- prod(nums)


```

# Day 4

https://adventofcode.com/2020/day/4

## Part 1
For day 4 we were given a messy data set with sets of variables with missing or present values (presented in an inconsistent delimitation). We had to determine how many sets were complete, not counting one particular variable (cid).

```{r}
day4 <- read_file("data/day4.txt")


day4 <- day4 %>% str_extract_all(regex(r'{(^.+$(\r\n)?)+}', multiline = TRUE))
day4 <- day4[[1]]
  

passport <- 1:length(day4)

day4_codes_list <- lapply(passport, function(iter){
    strsplit(day4[iter], ":|\r\n| ") %>% 
    unlist() %>% 
    matrix(ncol = 2, byrow = T) %>% 
    as_tibble() %>% 
    mutate(id = iter)
})

day4_codes <- do.call("rbind", day4_codes_list) %>% 
  pivot_wider(id_cols = id,
              names_from = V1,
              values_from = V2)

day4_valid <- day4_codes %>% 
  select(-cid) %>% 
  na.omit()

solution_day4_p1 <- dim(day4_valid)[1]

```

## Part 2

Given a set of conditions for each variable we had to determine not only whether each set was complete, but whether every variable was valid.

```{r}
day4_valid_2 <- day4_valid %>% 
  separate(hgt, into = c("hgt", "units"), sep = "(?<=[0-9])(?=[a-z])") %>% 
  na.omit() %>% 
  mutate(
    byr = as.numeric(byr),
    iyr = as.numeric(iyr),
    eyr = as.numeric(eyr),
    hgt = as.numeric(hgt)
  )

present <- 1:dim(day4_valid_2)[1]
num_valid <- lapply(present, function(iter){
  byr <- day4_valid_2$byr[iter] >= 1920 & day4_valid_2$byr[iter] <= 2002
  iyr <- day4_valid_2$iyr[iter] >= 2010 & day4_valid_2$iyr[iter] <= 2020
  eyr <- day4_valid_2$eyr[iter] >= 2020 & day4_valid_2$eyr[iter] <= 2030
  if(day4_valid_2$units[iter] == "cm"){
    hgt <- day4_valid_2$hgt[iter] >= 150 & day4_valid_2$hgt[iter] <= 193
  } else{if (day4_valid_2$units[iter] == "in"){
    hgt <- day4_valid_2$hgt[iter] >= 59 & day4_valid_2$hgt[iter] <= 76
  } else{
    hgt <- FALSE
  }
  }
  hcl <- grepl("#[0-9A-Fa-f]", day4_valid_2$hcl[iter])
  ecl <- day4_valid_2$ecl[iter] %in% c("amb", "blu", "brn", "gry", "grn", "hzl", "oth")
  pid <- grepl("^[0-9]{9}$", day4_valid_2$pid[iter])
  if (byr & iyr & eyr & hgt & hcl & ecl & pid){
    return (1)
  } else {
    return(0)
  }
})
solution_day4_p2 <- do.call("sum", num_valid)
```


# Day 5

## Part 1
For day 5, given a list of binary partitioned seats find the highest seat id. The first 7 characters gave the row (0 -127) where B's represented in the top half and F's in the bottom half. The last three characters gave the seat column (0 - 7) where R's were the top half and L's the bottom half. Seat id was calculated as row * 44 + seat.


```{r}
day5 <- read_csv("data/day5.txt", col_names = F)

day5 <- day5 %>% 
  separate(X1, 
           into = c("row", "seat"),
           sep = 7) %>% 
  rowid_to_column()
row_high <- day5[order(day5$row, -rank(day5$seat)),][[2]][1]
seat_high <- day5[order(day5$row, -rank(day5$seat)),][[3]][1]

row_high <- str_extract_all(row_high, boundary("character")) %>% 
  unlist()

seat_high <- str_extract_all(seat_high, boundary("character")) %>% 
  unlist()
  


row_number <- function(row){
  low <- 0
  high <- 127
  range <- 127
  for (i in 1:7){
    if (row[i] == "B"){
      low <- high - floor(range/2)
      high <- high
      } else{
        low <- low
        high <- low + floor(range/2)
        }
    range <- high - low
  }
  return(high)
}

row_number(row_high)

seat_number <- function(seat){
  low <- 0
  high <- 7
  range <- 7
  for(i in 1:3){
    if(seat[i] == "R"){
      low <- high - floor(range/2)
      high <- high
    } else{
      low <- low
      high <- low + floor(range/2)
    }
    range <- high - low
  }
  return(high)
}

solution_day5_p1 <- row_number(row_high)*8 + seat_number(seat_high)

```


## Part 2
For the second part we had to find the one missing seat. We only knew that the seat id -1 and +1 both existed.

```{r}

seat_ids <- lapply(day5$rowid, function(iter){
  row <- str_extract_all(day5$row[iter], boundary("character")) %>%  unlist()
  seat <- str_extract_all(day5$seat[iter], boundary("character")) %>%  unlist()
  return(c(day5$rowid[iter], row_number(row)*8 + seat_number(seat)))
})

seat_ids <- do.call("rbind", seat_ids) %>% 
  as_tibble()
colnames(seat_ids) <- c("rowid", "seat")

all_seats <- tibble(seat = min(seat_ids$seat):max(seat_ids$seat))

solution_day5_p2 <- left_join(all_seats, seat_ids) %>% 
  filter(is.na(rowid)) %>% 
  select(seat)


```
# Day 6

## Part 1
Day 6 has us taking answers from individuals in a variety of groups (separated by blank lines). We had to find how many unique questions were answered "yes" to by any group member and summing between groups.

```{r}
head(read_csv("data/day6.txt"))

day6 <- read_file("data/day6.txt")

day6 <- day6 %>% str_extract_all(regex(r"{(^.+$(\n)?)+}", multiline = TRUE)) %>% 
  unlist()


unique_answers <- lapply(day6, function(group){
  yeses <- unique(strsplit(group, "|\n")[[1]])
  yeses <- yeses[yeses != ""]
  return(length(yeses))
})

solution_day6_p1 <- do.call("sum", unique_answers)

```

## Part 2
For part 2 we instead had to find the number of questions in which <b>all</b> group members answered "yes" to.

```{r message=FALSE, warning=FALSE}

all_answers <- lapply(day6, function(group){
  group_size <- length(strsplit(group, "\n")[[1]])
  yeses <- strsplit(group, "|\n")[[1]] 
  yeses <- yeses[yeses != ""]
  yesperq <- yeses %>% 
    as_tibble() %>% 
    group_by(value) %>% 
    summarise(repeats = length(value)) %>% 
    filter(repeats == group_size)
  return(dim(yesperq)[1])
})
solution_day6_p2 <- do.call("sum", all_answers)

```



# Day 7

## Part 1

We were given a list of rules of where each type of bags contains a number of each other type of bag. For part 1 we needed to calculate how many types of bags would hold (directly and indirectly) a specific type of bag (shiny gold).

```{r}
day7 <- read_delim("data/day7.txt", delim = "\n", col_names = F)

day7 <- day7 %>% 
  separate(X1, into = c("container", "x1", "x2", "x3", "x4"), sep = " bags contain |, ", fill = "right") %>% 
  pivot_longer(cols = -container,
               names_to = NULL,
               values_to = "inside",
               values_drop_na = TRUE) %>% 
  separate(inside, into = c("number", "inside"), sep = " ", extra = "merge")

day7$inside <- gsub("\\.| bag| bags", "", day7$inside)



type <- "shiny gold"
all_bags <- vector()

while (length(type) > 0){
  holder <- vector()
  for (i in 1:length(type)){
    holder <- unique(c(holder, day7$container[day7$inside == type[i]]))
  }
  type <- holder[!(holder %in% all_bags)]
  all_bags <- unique(c(all_bags, holder))
}
 
solution_day7_p1 <- length(all_bags)

```


## Part 2

For part 2 we needed to find how many bags were within a particular type of bag (shiny gold).

```{r}
day7$number <- as.numeric(day7$number)
day7 <- na.omit(day7)

type <- tibble(number = 1, inside = "shiny gold")
all_bags <- tibble()

while (length(type$number) > 0){
  held <- tibble()
  for (i in 1:length(type$number)){
    held <- tibble(number = day7$number[day7$container == type$inside[i]], inside = day7$inside[day7$container == type$inside[i]]) %>% 
      mutate(number = number * type$number[i]) %>% 
      rbind(held)
  }
  type <- held
  all_bags <- rbind(all_bags, held)
}

solution_day7_p2 <- sum(all_bags$number)

```

# Day 8

## Part 1
Day 8 has us debugging a program in an infinite loop. There are three commands (acc, jmp, and nop). If acc is read then the accumulator (starting at 0) increases by the numeric value on that line and then the next line is read. If jmp is read, we advance the number of lines given (or go back if the number is negative). If nop is read nothing happens and the next line is read. The goal for part 1 is to find the accumulator value when a line is first read for the second time (i.e. the beginning of an infinite loop).


```{r}
day8 <- read_csv("data/day8.txt", col_names = F)
head(day8)

day8 <- day8 %>% 
  separate(X1, into = c("command", "step"), sep = " ") %>% 
  mutate(step = as.numeric(step),
         visit = 0)

accumulator <- 0
i <- 1
while(!(2 %in% day8$visit)){
  day8$visit[i] <- day8$visit[i] + 1
  if(day8$visit[i] < 2){
    if(day8$command[i] == "acc"){
      accumulator <- accumulator + day8$step[i]
      i <- i + 1}
    else{
      if (day8$command[i] == "jmp"){
        i <- i + day8$step[i]}
      else{
        if (day8$command[i] == "nop"){
          i <- i + 1}
      }
    }
  } 
}

solution_day8_p1 <- accumulator
```


## Part 2

To fix the problem we have to change either one jmp to a nop or a nop to a jmp. After finding which is correct we must return the final accumulator value when the program is run through.

```{r}
day8 <- day8 %>% rowid_to_column()

changes <- day8$rowid[day8$command %in% c("jmp", "nop") & day8$visit == 1]


test_run <- function(data){
   accumulator <- 0
   i <- 1
   complete <- TRUE
   while(complete & i <= dim(data)[1]){
     data$visit[i] <- data$visit[i] + 1
     if(data$visit[i] < 2){
       if(data$command[i] == "acc"){
         accumulator <- accumulator + data$step[i]
         i <- i + 1}
       else{
         if (data$command[i] == "jmp"){
           i <- i + data$step[i]}
         else{
           if (data$command[i] == "nop"){
             i <- i + 1}
         }
      }
     }else{
      complete <- FALSE
    }
   }
   return(tibble(complete = complete, acc = accumulator))
}



j <- 1
test <- tibble(complete = FALSE, acc = 0)
while(!test$complete){
  day8_test <- day8
  if(day8_test$command[day8$rowid == changes[j]] == "jmp"){
    day8_test$command[day8$rowid == changes[j]] <- "nop"
  }else{
    day8_test$command[day8$rowid == changes[j]] <- "jmp"
  }
  day8_test$visit <- 0
  
  test <- test_run(day8_test)
  j <- j + 1
}

solution_day8_p2 <- test$acc

```

# Day 9

## Part 1

We had a list of numbers. Starting at the 26th number, a number was considered valid if any two numbers in the previous 25 numbers added to that number. We had to find the first invalid number.

```{r}
day9 <- read_csv("data/day9.txt", col_names = FALSE)

day9 <- day9 %>% 
  transmute(numbers = X1)

i <- 1
k <- TRUE
while(k){
  preamble <- day9[i:(i+24),1]
  total <- day9$numbers[i+25]
  high_values <- preamble[preamble$numbers > total/2,]
  low_values <- preamble[preamble$numbers < total/2,]
  pair <- high_values %>% 
    mutate(search_value = total - numbers) %>% 
    inner_join(low_values, by = c("search_value" = "numbers"))
  
  if(dim(pair)[1] > 0){
    i <- i + 1
    
  }else{ if(length(preamble$numbers[preamble$numbers == day9$numbers[i +25]/2]) >= 2){
    i <- i + 1
  }else{
    
  }
    k <- FALSE
  }
}

solution_day9_p1 <- day9$numbers[i + 25]
```


## Part 2

We had to find the list of contiguous numbers (of at least 2) that summed to the invalid number found in part 1.

```{r}


i <- 1
j <- 1
k <- 0
cont <- vector()
while (k != solution_day9_p1 | length(cont) < 2){
  if(k < solution_day9_p1){
    cont[j] <- day9$numbers[i]
    k <- sum(cont)
    j <- j + 1
    i <- i + 1
  }else{
    cont <- cont[2:length(cont)]
    k <- sum(cont)
    j <- j - 1
  }
}

solution_day9_p2 <- min(cont) + max(cont)





```

# Day 10

## Part 1
We were given a list of numbers and needed to find the distance between each number (in order).

```{r}
day10 <- read_csv("data/day10.txt", col_names = F)

day10 <- day10 %>% 
  transmute(jolts = sort(X1))

day10 <- rbind(0,day10, max(day10 + 3)) %>% 
  rowid_to_column()

jolts <- lapply(day10$rowid[1:(dim(day10)[1]-1)], function(iter){
  return(day10$jolts[iter + 1] - day10$jolts[iter])
})

jolts <- do.call("rbind", jolts) 
  

jolt_summary <- jolts %>% 
  as_tibble() %>% 
  group_by(V1) %>% 
  summarise(count = n())


solution_day10_p1 <- jolt_summary$count[1] * jolt_summary$count[2]
```


## Part 2
With the same list, if each number can only move to a number within 3 of it, we needed to find the number of possible paths that would get from 0 to the maximum number +3.

```{r message=FALSE, warning=FALSE}
ends <- tibble(num = 1, value = 0)


for (iter in 1:(dim(day10)[1]-1)){
  if (((day10$jolts[iter + 1] - day10$jolts[iter]) == 3)|((day10$jolts[iter + 2] - day10$jolts[iter]) > 3)){
    add_path <- day10$jolts[iter + 1]
  }else{ if((day10$jolts[iter + 3] - day10$jolts[iter]) > 3){
    add_path <- day10$jolts[(iter+1):(iter + 2)]
  }else{
    add_path <- day10$jolts[(iter+1):(iter + 3)]
  }
  }
   ends <- rbind(ends, tibble(num = ends$num[ends$value == day10$jolts[iter]], value = add_path)) %>% 
     filter(value != day10$jolts[iter]) %>% 
     group_by(value) %>% 
     summarize(num = sum(num))
}

solution_day10_p2 <-ends$num
#print(ends$num, digits = 10)

```

# Day 11

## part 1

```{r}
day11 <- read_csv("data/day11.txt", col_names = F)

day11 <- day11 %>% 
  separate(col = X1, into = paste("x", 0:99, sep = ""), sep = "") %>% 
  select(-x0)



adjacent <- list(c(0,1),
                 c(0,-1),
                 c(1,-1),
                 c(1,0),
                 c(1,1),
                 c(-1,-1),
                 c(-1,0),
                 c(-1,1))

occupied <- function(delta, i, j){
  if (i + delta[1] < 1 | i + delta[1] > dim(day11)[1]){return(0)}
  if (j + delta[2] < 1 | j + delta[2] > dim(day11)[2]){return(0)}
  
  if(state_current[i + delta[1], j + delta[2]] == "#"){
    return(1)
  }else{
    return(0)
  }
}

sum.occupied <- function(out,input, i, j){
  res <- out + occupied(input, i,j)
  return(res)
}

#Using lapply
i_range <- 1:dim(day11)[1]
j_range <- 1:dim(day11)[2]
state_current <- day11
state_future <- day11
k <- FALSE


iter <- list()
m <- 1
for (i in i_range){
  for(j in j_range){
    iter[[m]] <- c(i,j)
    m <- m+1
  }
}


seat_state <- function(iter){
  if(state_current[iter[1], iter[2]] != "."){
        count_occupied <- reduce(adjacent, sum.occupied, i = iter[1], j = iter[2], .init = 0)
        if (state_current[iter[1], iter[2]] == "L"){
      if(count_occupied == 0){
        return("#")
      }else{
        return("L")
      }
      }
    if ((state_current[iter[1], iter[2]] == "#")){
      if(count_occupied >= 4){
        return("L")
      }else{
      return("#")
    }
  }
    }else{
    return(".")
  }
}


while(!k){
  state <- lapply(iter, seat_state)
  state_future <- unlist(state) %>% 
    matrix(nrow = length(i_range), ncol = length(j_range), byrow = TRUE)
  k <- identical(state_current, state_future)
  state_current <- state_future
}

solution_day11_p1 <- length(state_future[state_future == "#"])



```

## Part 2

```{r}

state_current <- day11
state_future <- day11
k <- FALSE

adjacent_array <- array(dim = c(length(i_range), length(j_range), 12), dimnames = list(NULL, NULL, c("right",
                                                                                                    "right.up.col",
                                                                                                    "right.up.row",
                                                                                                    "right.down.col",
                                                                                                    "right.down.row",
                                                                                                    "left",
                                                                                                    "left.up.col",
                                                                                                    "left.up.row",
                                                                                                    "left.down.col",
                                                                                                    "left.down.row",
                                                                                                    "up",
                                                                                                    "down")))
iter <- list()
m <- 1
for (i in i_range){
  for(j in j_range){
    iter[[m]] <- c(i,j)
    m <- m + 1
  }
}

for (i in seq_along(i_range)){
  for (j in seq_along(j_range)){
    #right
    k <- j + 1
    if (k <= max(j_range)){n <- state_current[i,k] == "."} else{n <- FALSE}
    while(n){
      k <- k + 1
      if (k <= max(j_range)){n <- state_current[i,k] == "."} else{n <- FALSE}
    }
    adjacent_array[i,j,1] <- if(k <= max(j_range)){k - j}else{NA}
    
    #right.up
    k <- j + 1
    l <- i - 1
    if (k <= max(j_range) & l > 0){n <- state_current[l,k] == "."} else{n <- FALSE}
    
    while(n){
      k <- k + 1
      l <- l - 1
      if (k <= max(j_range) & l > 0){n <- state_current[l,k] == "."} else{n <- FALSE}
    }
    adjacent_array[i,j,2] <- if(k <= max(j_range)){k - j}else{NA}
    adjacent_array[i,j,3] <- if(l > 0){l - i}else{NA}
    
    #right.down
    k <- j + 1
    l <- i + 1
    if (k <= max(j_range) & l <= max(i_range)){n <- state_current[l,k] == "."} else{n <- FALSE}
    while(n){
      k <- k + 1
      l <- l + 1
      if (k <= max(j_range) & l <= max(i_range)){n <- state_current[l,k] == "."} else{n <- FALSE}
      
    }
    adjacent_array[i,j,4] <- if(k <= max(j_range)){k - j}else{NA}
    adjacent_array[i,j,5] <- if(l <= max(i_range)){l - i}else{NA}
    
    #left
    k <- j - 1
    if (k > 0){n <- state_current[i,k] == "."} else{n <- FALSE}
    while(n){
      k <- k - 1
      if (k > 0){n <- state_current[i,k] == "."} else{n <- FALSE}
    }
    adjacent_array[i,j,6] <- if(k > 0){k - j}else{NA}
    
    #left.up
    k <- j - 1
    l <- i - 1
    if (k > 0 & l > 0){n <- state_current[l,k] == "."} else{n <- FALSE}
    while(n){
      k <- k - 1
      l <- l - 1
      if (k > 0 & l > 0){n <- state_current[l,k] == "."} else{n <- FALSE}
    }
    adjacent_array[i,j,7] <- if(k > 0){k - j}else{NA}
    adjacent_array[i,j,8] <- if(l > 0){l - i}else{NA}
    
    #left.down
    k <- j - 1
    l <- i + 1
    if (k > 0 & l <= max(i_range)){n <- state_current[l,k] == "."} else{n <- FALSE}
    while(n){
      k <- k - 1
      l <- l + 1
      if (k > 0 & l <= max(i_range)){n <- state_current[l,k] == "."} else{n <- FALSE}
      
    }
    adjacent_array[i,j,9] <- if(k > 0){k - j}else{NA}
    adjacent_array[i,j,10] <- if(l <= max(i_range)){l - i}else{NA}
    
    #up
    l <- i - 1
    
    
    if (l > 0){n <- state_current[l,j] == "."} else{n <- FALSE}
    while(n){
      l <- l - 1
      if (l > 0){n <- state_current[l,j] == "."} else{n <- FALSE}
    }
    adjacent_array[i,j,11] <- if(l > 0){l - i}else{NA}
    
    #down
    l <- i + 1
    if (l <= max(i_range)){n <- state_current[l,j] == "."} else{n <- FALSE}
    while(n){
      l <- l + 1
      if (l <= max(i_range)){n <- state_current[l,j] == "."} else{n <- FALSE}
    }
    adjacent_array[i,j,12] <- if(l <= max(i_range)){l - i}else{NA}
  }
}

occupied_p2 <- function(delta, i, j){
  if (is.na(delta[1])){return(0)}
  if (is.na(delta[2])){return(0)}
  
  if(state_current[i + delta[1], j + delta[2]] == "#"){
    return(1)
  }else{
    return(0)
  }
}

sum.occupied_p2 <- function(out,input, i, j){
  res <- out + occupied_p2(input, i,j)
  return(res)
}

seat_state_p2 <- function(iter){
  if(state_current[iter[1], iter[2]] != "."){
    
    adjacent <- list(c(0,adjacent_array[iter[1],iter[2],1]),
                 c(0,adjacent_array[iter[1],iter[2],6]),
                 c(adjacent_array[iter[1],iter[2],3], adjacent_array[iter[1],iter[2],2]),
                 c(adjacent_array[iter[1],iter[2],5], adjacent_array[iter[1],iter[2],4]),
                 c(adjacent_array[iter[1],iter[2],8], adjacent_array[iter[1],iter[2],7]),
                 c(adjacent_array[iter[1],iter[2],10], adjacent_array[iter[1],iter[2],9]),
                 c(adjacent_array[iter[1],iter[2],11], 0),
                 c(adjacent_array[iter[1],iter[2],12], 0))
    count_occupied <- reduce(adjacent, sum.occupied_p2, i = iter[1], j = iter[2], .init = 0)
    if (state_current[iter[1], iter[2]] == "L"){
      if(count_occupied == 0){
        return("#")
      }else{
        return("L")
      }
      }
    if ((state_current[iter[1], iter[2]] == "#")){
      if(count_occupied >= 5){
        return("L")
      }else{
      return("#")
    }
  }
    }else{
    return(".")
  }
}

ptm <- proc.time()
while(!k){
  state <- lapply(iter, seat_state_p2)
  state_future <- do.call("rbind",state) %>% 
    matrix(nrow = length(i_range), ncol = length(j_range), byrow = TRUE)
  k <- identical(state_current, state_future)
  state_current <- state_future
}
proc.time() - ptm

solution_day11_p2 <- length(state_current[state_current == "#"])

```

# Day 12

## Part 1

```{r}
day12 <- read_csv("data/day12.txt",col_names = FALSE)
head(day12)

day12 <- day12 %>% 
  separate(X1, into = c("direction", "distance"), sep = "(?<=[A-Z])(?=[0-9])") %>% 
  mutate(distance = as.numeric(distance))



move_func <- function(position, direction, distance){
  if(direction == "N"){
    position[1] <- position[1] + distance
    return(position)
  }
  if(direction == "S"){
    position[1] <- position[1] - distance
    return(position)
  }
  if(direction == "E"){
    position[2] <- position[2] + distance
    return(position)
  }
  if(direction == "W"){
    position[2] <- position[2] - distance
    return(position)
  }
}

rotate_func <- function(face, direction, angle){
  if(direction == "R"){
    return(cardinals[which(cardinals == face)[1] + angle/90])
  }
  if(direction == "L"){
    return(cardinals[which(cardinals == face)[2] - angle/90])
  }
}

current_position <- c(0,0)
dir_face <- "E"
cardinals <- rep(c("N", "E", "S", "W"), 2)

for(i in 1:dim(day12)[1]){
   if(day12$direction[i] == "F"){
      current_position <- move_func(current_position, dir_face, day12$distance[i])
   }else{
     if (day12$direction[i] %in% c("N", "S", "E", "W")){
       current_position <- move_func(current_position, day12$direction[i], day12$distance[i])
     }else{
      dir_face <- rotate_func(dir_face, day12$direction[i], day12$distance[i])
    }
   }
}

solution_day12_p1 <- sum(abs(current_position))

```

## Part 2

```{r}
rotate_wp_func <- function(position, direction, angle){
  x <- position[1]
  y <- position[2]
  if(angle == 180){
    return(c(-x, -y))
  }
  if((direction == "R" & angle == 90)|(direction == "L" & angle == 270)){
    return(c(-y, x))
  }
  if((direction == "L" & angle == 90)|(direction == "R" & angle == 270)){
    return(c(y, -x))
  }
}

current_position <- c(0,0)
wp_position <- c(1, 10)


for(i in 1:dim(day12)[1]){
   if(day12$direction[i] == "F"){
      current_position <- current_position + wp_position * day12$distance[i]
   }else{
     if (day12$direction[i] %in% c("N", "S", "E", "W")){
       wp_position <- move_func(wp_position, day12$direction[i], day12$distance[i])
     }else{
      wp_position <- rotate_wp_func(wp_position, day12$direction[i], day12$distance[i])
     }
   }
  #print(list(progress = i, position = current_position, wp = wp_position))
}

solution_day12_p2 <- sum(abs(current_position))
```



# Day 13

## Part 1
```{r}
day13 <- read_file("data/day13.txt")

day13 <- strsplit(day13, split = "\n|,") %>% unlist()

value <- day13[1] %>% as.numeric()

numbers <- day13[-1]
numbers <- numbers[numbers != "x"] %>% as.numeric()

multiples <- value/numbers 
multiples <- lapply(multiples, ceiling) %>% unlist()
differentials <- multiples * numbers - value
solution_day13_p1 <- numbers[differentials == min(differentials)] * min(differentials)
```

## Part 2

```{r}
day13 <- day13[-1] %>%
  tibble() %>%
  rowid_to_column() %>%
  rename(bus = ".") %>%
  filter(bus != "x") %>%
  transmute(bus = as.numeric(bus),
         remainder = 1-rowid)

#day13 <- day13[order(-day13$bus),]

day13 <- c(17,"x",13,19) %>% 
  tibble() %>% 
  rowid_to_column() %>% 
  rename(bus = ".") %>%
  filter(bus != "x") %>%
  transmute(bus = as.numeric(bus),
         remainder = 1-rowid)

day13 <- c(67,7,59,61) %>% 
  tibble() %>% 
  rowid_to_column() %>% 
  rename(bus = ".") %>%
  filter(bus != "x") %>%
  transmute(bus = as.numeric(bus),
         remainder = 1 - rowid)



gcd_func <- function(r1, r2){
  s1 <- 1
  s2 <- 0
  t1 <- 0
  t2 <- 1
  r <- 1
  while(r > 0){
    q <- r1 %/% r2
    r <- r1 %% r2
    s <- s1 - q * s2
    t <- t1 - q * t2
    if(r > 0){
      r1 <- r2
      r2 <- r
      s1 <- s2
      s2 <- s
      t1 <- t2
      t2 <- t
    }
  }
  return(list(gcd = r2, coef1 = s2, coef2 = t2))
}
  
iter <- 2:dim(day13)[1]
a <- as.bigz(day13$bus[1])
remainder <- as.bigz(day13$remainder[1])

for(i in iter){
  b <- as.bigz(day13$bus[i])
  gcds <- gcd_func(a,b)
  x <- as.bigz(day13$remainder[i]) * a * as.bigz(gcds$coef1) + remainder * b * as.bigz(gcds$coef2)
  remainder <- x %% (a*b)
  a <- a * b
}

```

